Think deep for reef resilience

### 2025-08-17 paper/paper_draft.qmd
Standardise cluster language
Ensure that "cluster" is used when talking about time series clustering allocated categories, and associated analyses that revolve around these low, medium, high clusters.

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High categories. These characteristics include reef median depth,
mean time series DHW, Pearson’s correlation between total coral cover and inverse DHW time
series, proportion of coral habitable area occupied by corals in the first timestep, ratio
of source-to-sink larval connectivity and total strength of all larval connections between
reefs.

@@

Of the GCMs explored, NorESM2-MM had the lowest Equilibrium Climate Sensitivity (2.5°C) and
reefs exhibit continuing recovery and decline dynamics up to 2075 under this hypothetical
climate future. Unsurprisingly, recovery phases were predominantly seen in the reefs in the
High category, with Medium and Low exhibiting lesser probability of such periods of
recovery. Generally, recovery and increases in coral cover were visible in simulated results
up to 2045. Under NorESM2-MM, the Far Northern Management area were found to have the most
similar temporal behaviour between Low and High categories compared to other Management
areas (@tbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
category to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 (@fig-methods-dhws). Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
(@tbl-cluster-cid).

@@

Comparing across GCMs, it is common for reefs to switch between categories, including from
Low to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2429 reefs, 20%
(503) of reefs have consistent category assignment across all GCMs. Of the 503 consistently
assigned reefs, 49% are within the Far Northern Management area, with 7%, 13% and 31%
occurring in the Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management
areas respectively (@sfig-gcm-cluster-map). Of the 503 consistently assigned reefs, 32%, 7%
and 61% are clustered into Low, Medium and High clusters respectively. High levels of
assignment variability across GCMs are expected as due to spatial differences in each given
GCM’s DHW trajectories.

@@

In contrast to the broad temporal behaviour at the Management area scale, the relationship
between reef characteristics and assigned categories were found to be similar at the
bioregion scale across GCMs. Results are only shown and discussed for ACCESS-ESM1-5, an
intermediate ECS scenario (@sfig-ecs). Those interested in results for the other assessed
GCMs are referred to Grier and Iwanaga [-@grier2025]. Across bioregions, reefs in the High
category not only had greater levels of coral recovery but displayed persistence to 2050
(@sfig-access-esm1-5-bioregion-timeseries). Reefs in the High category were consistently
found to have the greatest median reef depths across bioregions
(@fig-access-esm1-5-bioregion-depth).

@@

In many bioregions reefs in the Low category have the lowest correlation between coral cover
and inverse-DHW time series whereas reefs in the High category were found to have the
strongest correlation (@sfig-access-esm1-5-bioregion-dhwcor). Across six of the 29
bioregions reef mean DHW levels increased as cluster cover decreased
(@sfig-access-esm1-5-bioregion-dhwmean). There were no consistent patterns across bioregions
in their reefs' initial proportion of habitable area occupied by corals and cluster
assignment (@sfig-access-esm1-5-bioregion-initialproportion). Across most bioregions there
were not consistent differences between clusters in connectivity metrics such as Log-10
source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10 total
connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity).

@@

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and more stable, assignment
probability across depth ranges (@fig-ts-cluster-pdp; Panel B). A large increase in the
probability of a reef being categorised as High can be seen between 10 and 15 meters depth.
This is at the edge of the depth range used in current deployment efforts [@howlett2021;
@randall2021; @whitman2024] but indicates a depth range that could be adopted to maximise
longevity and benefits of, for example, coral seeding activities. Across mean DHW values,
assignment probability is roughly equal across the three categories up to 6 DHW. A sharp
delineation can be seen \> 6 DHW when comparing High with other categories, with lower
assignment probability (\~25%) compared to Medium and Low (\~40%) (@fig-ts-cluster-pdp;
Panel B).

@@

![Feature importance scores (Panel A) of the Random Forest classifier, and Partial
Dependence Plots (PDPs) for reef characteristics of interest indicating the probability of a
reef being classified to one of the three categories (Panel B). Partial Dependence Plots for
bioregions were excluded as they were found to be uninformative. PDPs indicate the
probability of a reef from the test dataset being clustered into each cluster category
depending on the predictor value on the x axis, when adjusted for values of other
predictors. Green, orange and blue lines indicate cluster assignment probability for Low,
Medium and High clusters respectively. Transparent grey points above probability lines
represent the predictor values present in the test dataset, providing a rough indication of
predictor value distribution across
reefs.](../figs/ts_cluster_rf_feature_analysis.png){#fig-ts-cluster-pdp fig-align="center"
width="14.82cm"}

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable restoration
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, restoration depth is likely constrained by
snorkelling and diving deployment and monitoring logistical accessibility, visual
identification of sites from the surface, in combination with tourism return value. Our
results suggest that of the assessed reefs occurring within the 1-10m depth range (1736
reefs), 62% (1082) of reefs fall into the low and medium clusters (across all GCMs at the
GBRMPA management region scale), with many unable to maintain positive carbonate budgets
under future conditions. Partial dependence analysis revealed a range of depths from 10-15m
where the probability of a reef being clustered into the High cover category increased
substatially. Considering this 10-15m depth range when assessing possible restoration sites
may reduce the exposure of planted and seeded corals to high thermal stress, while still
maintaining effective logistical access to sites. Reef intervention planning is a complex
and challenging decision-making process [@anthony2020], and our results suggest that
deployment site depth may be an important factor to additionally consider when assessing the
long-term resilience of sites.

@@

The CID timeseries clustering metric corrects the Euclidean distance between pairs of time
series (which accounts for magnitude differences) with the complexity of each time series
(obtained through a measure of their variability) to ensure reefs are clustered based on
their underlying temporal dynamics rather than simply their average coral cover. The process
creates a distance matrix of n \* n dimensions (where n is the number of time series in the
group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The
three reef clusters are labelled with qualitative-ordered categories from Low, Medium and
High based on their median cover relative to each other cluster over the 2030-2060 period.
The selected time frame prevents the classification from being overly influenced by any
steep declines in the initial years of simulations and avoids the post-2060 time period
where reefs are projected to be under severe stress and in a declined state [@bozec2025].

>>>>

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High clusters. These characteristics include reef median depth,
mean time series DHW, Pearson’s correlation between total coral cover and inverse DHW time
series, proportion of coral habitable area occupied by corals in the first timestep, ratio
of source-to-sink larval connectivity and total strength of all larval connections between
reefs.

@@

Of the GCMs explored, NorESM2-MM had the lowest Equilibrium Climate Sensitivity (2.5°C) and
reefs exhibit continuing recovery and decline dynamics up to 2075 under this hypothetical
climate future. Unsurprisingly, recovery phases were predominantly seen in the reefs in the
High cluster, with Medium and Low exhibiting lesser probability of such periods of
recovery. Generally, recovery and increases in coral cover were visible in simulated results
up to 2045. Under NorESM2-MM, the Far Northern Management area were found to have the most
similar temporal behaviour between Low and High clusters compared to other Management
areas (@tbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
cluster to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 (@fig-methods-dhws). Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
(@tbl-cluster-cid).

@@

Comparing across GCMs, it is common for reefs to switch between clusters, including from
Low to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2429 reefs, 20%
(503) of reefs have consistent cluster assignment across all GCMs. Of the 503 consistently
assigned reefs, 49% are within the Far Northern Management area, with 7%, 13% and 31%
occurring in the Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management
areas respectively (@sfig-gcm-cluster-map). Of the 503 consistently assigned reefs, 32%, 7%
and 61% are clustered into Low, Medium and High clusters respectively. High levels of
assignment variability across GCMs are expected as due to spatial differences in each given
GCM’s DHW trajectories.

@@

In contrast to the broad temporal behaviour at the Management area scale, the relationship
between reef characteristics and assigned clusters were found to be similar at the
bioregion scale across GCMs. Results are only shown and discussed for ACCESS-ESM1-5, an
intermediate ECS scenario (@sfig-ecs). Those interested in results for the other assessed
GCMs are referred to Grier and Iwanaga [-@grier2025]. Across bioregions, reefs in the High
cluster not only had greater levels of coral recovery but displayed persistence to 2050
(@sfig-access-esm1-5-bioregion-timeseries). Reefs in the High cluster were consistently
found to have the greatest median reef depths across bioregions
(@fig-access-esm1-5-bioregion-depth).

@@

In many bioregions reefs in the Low cluster have the lowest correlation between coral cover
and inverse-DHW time series whereas reefs in the High cluster were found to have the
strongest correlation (@sfig-access-esm1-5-bioregion-dhwcor). Across six of the 29
bioregions reef mean DHW levels increased as cluster cover decreased
(@sfig-access-esm1-5-bioregion-dhwmean). There were no consistent patterns across bioregions
in their reefs' initial proportion of habitable area occupied by corals and cluster
assignment (@sfig-access-esm1-5-bioregion-initialproportion). Across most bioregions there
were not consistent differences between clusters in connectivity metrics such as Log-10
source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10 total
connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity).

@@

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and more stable, assignment
probability across depth ranges (@fig-ts-cluster-pdp; Panel B). A large increase in the
probability of a reef being categorised as High can be seen between 10 and 15 meters depth.
This is at the edge of the depth range used in current deployment efforts [@howlett2021;
@randall2021; @whitman2024] but indicates a depth range that could be adopted to maximise
longevity and benefits of, for example, coral seeding activities. Across mean DHW values,
assignment probability is roughly equal across the three clusters up to 6 DHW. A sharp
delineation can be seen \> 6 DHW when comparing High with other clusters, with lower
assignment probability (\~25%) compared to Medium and Low (\~40%) (@fig-ts-cluster-pdp;
Panel B).

@@

![Feature importance scores (Panel A) of the Random Forest classifier, and Partial
Dependence Plots (PDPs) for reef characteristics of interest indicating the probability of a
reef being classified to one of the three clusters (Panel B). Partial Dependence Plots for
bioregions were excluded as they were found to be uninformative. PDPs indicate the
probability of a reef from the test dataset being clustered into each category
depending on the predictor value on the x axis, when adjusted for values of other
predictors. Green, orange and blue lines indicate cluster assignment probability for Low,
Medium and High clusters respectively. Transparent grey points above probability lines
represent the predictor values present in the test dataset, providing a rough indication of
predictor value distribution across
reefs.](../figs/ts_cluster_rf_feature_analysis.png){#fig-ts-cluster-pdp fig-align="center"
width="14.82cm"}

@@

Current reef restoration and intervention-investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable restoration
sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are
no explicit justifications for this depth range, restoration depth is likely constrained by
snorkelling and diving deployment and monitoring logistical accessibility, visual
identification of sites from the surface, in combination with tourism return value. Our
results suggest that of the assessed reefs occurring within the 1-10m depth range (1736
reefs), 62% (1082) of reefs fall into the low and medium clusters (across all GCMs at the
GBRMPA management region scale), with many unable to maintain positive carbonate budgets
under future conditions. Partial dependence analysis revealed a range of depths from 10-15m
where the probability of a reef being clustered into the High cover cluster increased
substatially. Considering this 10-15m depth range when assessing possible restoration sites
may reduce the exposure of planted and seeded corals to high thermal stress, while still
maintaining effective logistical access to sites. Reef intervention planning is a complex
and challenging decision-making process [@anthony2020], and our results suggest that
deployment site depth may be an important factor to additionally consider when assessing the
long-term resilience of sites.

@@

The CID timeseries clustering metric corrects the Euclidean distance between pairs of time
series (which accounts for magnitude differences) with the complexity of each time series
(obtained through a measure of their variability) to ensure reefs are clustered based on
their underlying temporal dynamics rather than simply their average coral cover. The process
creates a distance matrix of n \* n dimensions (where n is the number of time series in the
group) that is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The
three reef clusters are labelled with qualitative-ordered clusters from Low, Medium and
High based on their median cover relative to each other cluster over the 2030-2060 period.
The selected time frame prevents the classification from being overly influenced by any
steep declines in the initial years of simulations and avoids the post-2060 time period
where reefs are projected to be under severe stress and in a declined state [@bozec2025].
